---
title: "Immigration Detention Centers Analysis"
author: "Michael Ale"
date: "2025-04-26"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Importing Necessary Libraries

```{r}
#tinytex::install_tinytex(force = TRUE)
library(tidyverse)
library(dplyr)
```

# 2. Loading the Dataset

```{r}
df <- read.csv("immigration-detention.csv")
head(df)
```
# 3. Cleaning the Dataset

### 3.1 Removing Headers in First Few Rows
```{r}
df <- read.csv("immigration-detention.csv", skip = 6)
head(df)
colnames(df)
dim(df)
```
### 3.2 Handling Special characters in Name
```{r}
df$Name<- iconv(df$Name,from = "",to = "UTF-8", sub = "")
df$Name <- gsub("[^A-Za-z ]", "", df$Name)#this ensures only A to Z and a to z character is in the dataframe
head(df$Name)
```
### 3.3 Handling Blanks in Name,City and State

```{r}
# Name
df <- df %>% mutate(Name = if_else(Name == "", "Unknown", Name))

# City
df$City <- trimws(df$City)
df <- df %>%
  mutate(City = if_else(Name == "GEAUGA COUNTY JAIL" & City == "", "CHARDON", City))

# State
df$State <- trimws(df$State)
df <- df %>%
  mutate(State = if_else(City == "ATLANTA", "GA", State),
         State = if_else(City == "ENCINAL", "TX", State))  # <-- Corrected to TX

```
### 3.4 Handling Dates
```{r}
df <- df %>%
  mutate(Last.Inspection.End.Date = as.Date(as.numeric(Last.Inspection.End.Date), origin = "1899-12-30"))
```
```{r}
# Fix Nevada separately
df <- df %>%
  mutate(Last.Inspection.End.Date = if_else(
    Name == "NEVADA SOUTHERN DETENTION CENTER",
    as.Date("2024-09-19"),
    Last.Inspection.End.Date
  ))
```
# 4.Analyze the Data

### 4.1 Create Total Population Column
```{r}
df <- df %>%
  mutate(Total.Population = Level.A + Level.B + Level.C + Level.D)

```
### 4.2 Top 10 Facilities by Population
```{r}
top_10_facilities <- df %>%
  arrange(desc(Total.Population)) %>%
  slice_head(n = 10)

```
### 4.3 Top 10 States by Number of Detention Centers
```{r}
Each_state_detention_center <- df %>%
  group_by(State) %>%
  summarise(Detention_center_count = n()) %>%
  arrange(desc(Detention_center_count)) %>%
  slice_head(n = 10)

```
### 4.4 Top 10 States By Population in Detention Centers
```{r}
State_total_population <- df %>%
  group_by(State) %>%
  summarise(Total = sum(Total.Population, na.rm = TRUE)) %>%
  arrange(desc(Total)) %>%
  slice_head(n = 10)

```
# 5. Visualization

### 5.1 Top 10 Largest Detention Facilities by Population
```{r}
ggplot(data = top_10_facilities, mapping =  aes(x = reorder(Name, Total.Population), y = Total.Population)) +
  geom_bar(stat = "identity", fill = "black") +
  geom_text(aes(label = round(Total.Population)), hjust = -0.5, size = 3) +
  coord_flip() +
  labs(title = "Top 10 Detention Facilities by Total Population",
       x = "Detention Facility",
       y = "Total Population") +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white"),#while saving image i wasn't getting clear plotting so fixing it white solve the issue
    plot.background = element_rect(fill = "white")
  )

```



### 5.2 Top 10 States by Number of Detention Centers
```{r}
ggplot(data = Each_state_detention_center,mapping =  aes(x = reorder(State, Detention_center_count), y = Detention_center_count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = Detention_center_count), hjust = -0.2, size = 3) + #adding text on each bar
  coord_flip() + #it flips the axes as when names were not clear.
  labs(title = "Top 10 States With Highest Number Of Detention centers",
       x = "State",
       y = "Number Of Detention Centers") +
  theme_minimal()+
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white"),#while saving image i wasn't getting clear plotting so fixing it white solve the issue
    plot.background = element_rect(fill = "white")
  )
```

### 5.3 Top 10 States by Population in Detention Centers
```{r}
ggplot(data = State_total_population, aes(x = reorder(State, Total), y = Total)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = round(Total)), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(title = "Top 10 States With Highest Population in Detention Centers",
       x = "State",
       y = "Total Population in Detention Centers") +
  theme_minimal()

```

